|

### 2021-11-11 submission/manuscript.rmd
update discussion

Fourth, we used *restricted cubic splines* (RCS) to relax the linearity
assumption on the effect of the linear predictor [@Harrell1988]. We considered
splines with 3, 4 and 5 knots to compare models with different levels of
flexibility.

@@

Increasing the sample size to `r 17000` favored the RCS-3 approach the
most, as it achieved lowest or close to lowest RMSE across all
scenarios (Figure \ref{fig:rmsesamplesize}). Especially in cases of
strong quadratic and non-monotonic deviations RCS-3 had lower error
rates
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms)
compared to the linear interaction approach
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of treatment-related harms strength. The
issues with the large error variability of the adaptive approach
improved with larger sample sizes.

@@

When we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = `r 4250`) the constant effect model outperformed all other
methods in the case of true constant treatment effects, but proved to
be the least robust to deviations. Again, RCS-3 had the lowest
error rates in the case of strong quadratic or monotonic deviations
and very comparable performance to the best-performing linear
interaction model in the case of strong linear deviations
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). The adaptive approach, though it
performed similar to the best performing method in each scenario, on
average, there was increased variability in error rates in the case of
strong linear and non-monotonic deviations. In these scenarios, the adaptive approach often selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
in the strong linear and non-monotonic deviation scenarios without
treatment-related harms, respectively).

@@

In terms of discrimination for benefit (OR = 0.8, N = `r 4250` and AUC
= 0.75), all methods performed similarly, on average (Figure
\ref{fig:discrimination}). In the case of non-monotonic deviations,
the constant effect model had much lower discriminative performance
compared to the rest of the methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model compared to the best-performing RCS-3
with
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
The linear interaction model was most stable in compared to the other
methods in terms of error variability. With increasing number of RCS
knots, we observed decreasing median values and increasing variability
of the c-for-benefit in all scenarios.

@@

As for calibration, the constant effects model outperformed all other
models in the case of true constant treatment effects, but it was
miscalibrated for all deviation scenarios (Figure
\ref{fig:calibration}). The linear interaction model showed best or
close to best calibration across all scenarios and only showed worse
calibration compared to RCS-3 in the case of non-monotonic deviations
and treatment-related harms. The adaptive approach was worse
calibrated in scenarios with strong linear and non-monotonic
deviations compared to the linear interaction model and RCS-3.

@@

The linear interaction model proved to be superior to alternative approaches
when predicting risk-based treatment benefit under a wide range of scenarios. It
generally had lower mean squared error, lower e-for-benefit and higher
c-for-benefit with lower variability across simulation replications. Models
including restricted cubic splines with 3 knots only outperformed the linear
interaction model in the presence of strong quadratic deviations from a constant
relative treatment effect.

Models including restricted cubic splines with 4 or 5 knots proved to be too
flexible, as indicated by higher RMSE, increased variability of discrimination
for benefit and worse calibration of benefit predictions. Even with larger
sample sizes and strong quadratic deviations from the base case scenario of
constant relative treatment effects, these more flexible restricted cubic
splines did not outperform the simpler RCS with 3 knots. These approaches may
only be helpful if we expect more extreme patterns of heterogeneous treatment
effects compared to the quadratic deviations considered here.

@@

The adaptive approach had adequate performance, following closely the
performance of the "true" model in most scenarios. However, with smaller sample
sizes it tended to "miss" the treatment-risk interactions and selected simpler
models (Supplementary Table S7). This resulted in increased RMSE variability in
these scenarios, especially in the case of true strong linear deviations from
the base case scenario. This tendency was mitigated when the discriminative
ability of the prediction model was higher. Therefore, in the case of smaller
sample sizes and/or poorly discriminating prediction models, the simpler linear
interaction model is a safer choice for predicting absolute benefits.

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation scenarios.
Nevertheless, our main conclusions did not change when we generated individual
treatment-covariate interactions (supplementary table/figure x). Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.

@@

In conclusion, when comparing different risk-based approaches to predicting 
individualized treatment benefit, a model including a linear treatment 
interaction with the prognostic index performed best in a wide range of 
scenarios. More flexible models with restricted cubic splines required larger
sample sizes and higher AUC of the prognostic index to outperform the linear
interaction model. An adaptive approach, selecting the model with the optimal 
AIC, had comparable performance to the best performing approach in most of the 
scenarios.

>>>>

Fourth, we used *restricted cubic splines* (RCS) to relax the
linearity assumption on the effect of the linear predictor
[@Harrell1988]. We considered splines with 3 (RCS-3), 4 (RCS-4) and 5
(RCS-5) knots to compare models with different levels of flexibility.

@@

Increasing the sample size to `r 17000` favored RCS-3 the most, as it
achieved lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower error rates (median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of treatment-related harms strength. The
issues with the large error variability of the adaptive approach
improved with larger sample sizes.

@@

When we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = `r 4250`) the constant effect model outperformed all other
methods in the case of true constant treatment effects, but proved to
be the least robust to deviations. Again, RCS-3 had the lowest
error rates in the case of strong quadratic or monotonic deviations
and very comparable performance to the best-performing linear
interaction model in the case of strong linear deviations
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). The adaptive approach, though it
performed similar to the best performing method in each scenario, on
average, had increased variability in error rates in the case of
strong linear and non-monotonic deviations. In these scenarios, the
adaptive approach often selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
in the strong linear and non-monotonic deviation scenarios without
treatment-related harms, respectively).

@@

In terms of discrimination for benefit (OR = 0.8, N = `r 4250` and AUC
= 0.75), all methods performed similarly, on average (Figure
\ref{fig:discrimination}). In the case of non-monotonic deviations,
the constant effect model had much lower discriminative performance
compared to the rest of the methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model compared to the best-performing RCS-3
with
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
The linear interaction model was the most stable compared to the other
methods in terms of error variability. With increasing number of RCS
knots, we observed decreasing median values and increasing variability
of the c-for-benefit in all scenarios.

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the case of true constant treatment
effects, but was miscalibrated for all deviation scenarios (Figure
\ref{fig:calibration}). The linear interaction model showed best or
close to best calibration across all scenarios and only showed worse
calibration compared to RCS-3 in the case of non-monotonic deviations
and treatment-related harms. The adaptive approach was worse
calibrated in scenarios with strong linear and non-monotonic
deviations compared to the linear interaction model and RCS-3.

@@

The linear interaction model displayed very good performance overall
under many of the considered simulation scenarios. Especially in cases
with smaller sample sizes and moderately performing baseline risk
prediction models it had lower RMSE, was better calibrated for benefit
and had better discrimination for benefit, even in scenarios with
strong quadratic deviations. However, in scenarios with true
non-monotonic deviations, the linear interaction model was
outperformed by RCS-3, especially in the presence of true
treatment-related harms. Increasing the sample size or the prediction
model's discrimination favored RCS-3 which had better or very
comparable performance to the linear interaction model, but was more
robust to non-monotonic deviations and the presence of
treatment-related harms.

RCS-4 and RCS-5 proved to be too flexible, as indicated by higher
RMSE, increased variability of discrimination for benefit and worse
calibration of benefit predictions. Even with larger sample sizes and
strong quadratic or non-monotonic deviations from the base case
scenario of constant relative treatment effects, these more flexible
restricted cubic splines did not outperform the simpler RCS-3 These
approaches may only be helpful if we expect more extreme patterns of
heterogeneous treatment effects compared to the quadratic deviations
considered here.

@@

The adaptive approach had adequate performance, following closely on
average the performance of the "true" model in most
scenarios. However, with smaller sample sizes it tended to "miss" the
treatment-risk interactions and selected simpler models (Supplementary
Table S7). This resulted in increased RMSE variability in these
scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario.  Therefore, in
the case of smaller sample sizes the simpler linear interaction model
is a safer choice for predicting absolute benefits.

Risk-based approaches to predictive HTE estimate treatment benefit as
a function of baseline risk. A limitation of our study is that we
assumed treatment benefit to be a function of baseline risk in the
majority of the simulation scenarios. We attempted to address that by
introducing constant moderate and strong treatment-related harms,
applied on the absolute scale. Also, we considered a small number of
scenarios with true treatment-covariate interactions, in which our
main conclusions remained the same (Supplement, XX). Future simulation
studies could explore the effect of more extensive deviations from
risk-based treatment effects.

@@

In conclusion, the best option for predicting individualized treatment
benefit using a risk-based approach depends on the setting. With
smaller sample sizes and/or moderately performing risk prediction
models the linear interaction approach is a viable option. When those
constraints are not present or when we anticipate non-negligible
treatment-related harms, RCS-3 is a better option in terms of error
rates, discrimination and calibration for benefit. With larger sample
size, an adaptive approach based on AIC can also be considered as a
more automated alternative.
